---
title: "R Notebook"
output: html_notebook
---

Inspired by [this](https://www.futilitycloset.com/2012/10/17/the-paradox-of-the-second-ace-2/)

## packages

```{r}
library(tidyverse)
```


## a deck of cards

```{r}
ranks=c("A","2","3","4","5","6","7","8","9","T","J","Q","K")
suits=c("S","H","D","C")
deck=crossing(rank=ranks,suit=suits)
deck
```

set random number seed

```{r}
set.seed(457299)
```


select 13 cards

```{r}
deck %>% sample_n(13) -> hand
```

make a function of this

```{r}
deal_hand=function(deck,cards=13) {
  deck %>% sample_n(cards)
}
```


## does a hand contain an ace, the ace of spades, more than one ace?

```{r}
count_aces=function(hand) {
  hand %>% filter(rank=="A") %>% nrow()
}
count_ace_spades=function(hand) {
  hand %>% filter(rank=="A", suit=="S") %>% nrow()
}
count_aces(hand)
count_ace_spades(hand)
```

sample many of these

```{r}
rerun(10000, deal_hand(deck)) %>% bind_rows(.id="hand_number") %>% nest(-hand_number, .key="hand") %>% 
  mutate(ace_count=map_int(hand,~count_aces(.))) %>% 
  mutate(spade_ace_count=map_int(hand,~count_ace_spades(.))) -> d
```

```{r}
d
```

Out of the times when there is at least one ace, how many times is there more than one?


```{r}
d %>% filter(ace_count>=1) %>% count(ace_count) %>% mutate(prop=n/sum(n))
```

```{r}
d %>% filter(spade_ace_count==1) %>% count(ace_count) %>% mutate(prop=n/sum(n))
```


number of aces held is hypergeometric:

```{r}
p1=dhyper(0:4,4,48,13)
p1
```

if she holds the spade ace, there are 3 aces remaining among the other 51 cards, so distribution of number of further aces held is

```{r}
p2=dhyper(0:3,3,48,12)
```

p(no further aces | spade ace):

```{r}
p2[1]/sum(p2)
```

out of the ways of holding at least one ace, how many involve holding no further aces?

```{r}
p1[2]/sum(p1[2:5])
```

or said differently:

```{r}
(tibble(x=0:4) %>% mutate(prob=dhyper(x,4,48,13)) -> p1)
```

and

```{r}
(tibble(x=0:3)) %>% mutate(prob=dhyper(x,3,48,12) -> p2)
```

distribution of total aces | at least one:

```{r}
p1 %>% filter(x>=1) %>% mutate(prob2=prob/sum(prob))
```

this reminds me of, in a 2-child family:

P(both girls | first child is a girl) = 1/2
P(both girls | at least one child is a girl) = 1/3 (only 1 possibility, BB, eliminated)