---
title: Odd sums
author: Ken
date: '2017-05-31'
slug: odd-sums
categories:
  - math
tags:
  - R
---

While waiting for my coffee at work this morning, I was leafing through a recreational mathematics journal. It said, "the numbers 1--9 are arranged at random in a 3 by 3 matrix. What is the probability that the row sums and column sums are all odd?"

Well, I thought, that ought to be easy enough to simulate. Randomly shuffle the numbers 1--9 using `sample`, and then insert them into a matrix with three columns. Calculate the row and column sums using `apply`, glue them together into one long vector, and test whether all of the values have remainder 1 when divided by 2 (a fancy way of saying "odd"). That led to this function:

```{r}
isodd=function(isprint=F) {
  x=sample(1:9)
  m=matrix(x,ncol=3)
  if (isprint) print(m)
  rowsums=apply(m,1,sum)
  colsums=apply(m,2,sum)
  v=(c(rowsums,colsums) %% 2)
  all(v==1)
}
```

I also put in an argument `isprint` that, if true, displays the randomly-generated matrix of shuffled values (for use when checking). So, let's see how it works, setting the random number seed for reproducibility. I decided to keep going until I get a matrix with all row and column sums odd, having previously convinced myself that this would happen eventually. As it happened, I didn't have to wait long:

```{r}
set.seed(3)
v=F
i=0
while (!v) {
  i=i+1
  print(i)
  v=isodd(T)
  print(v)
}
```

Most of the time, one (or more) of the rows and columns adds up to something even. For example, in the first simulated matrix, the second row adds up to 16 (and the first row adds up to 12). In the second simulated matrix, though, the rows add up to 21, 9 and 15, and the columns add up to 15, 19 and 11: all odd. So it is possible.

The next question is "how likely is it"? The standard simulation answer is "do it a lot of times, and see how often it happens for you". R offers `replicate` which will do exactly that:

```{r}
ans=replicate(10000,isodd())
table(ans)
```

So, all the rows and columns have odd sums just over 7% of the time.

The challenge in the place where I got the problem (Crux Mathematicorum) was to obtain the exact answer and prove that it is correct. 

If you look at the second example matrix above, the one with all row and column sums odd, there is a row with all odd numbers and a column with all odd numbers as well. Or, to put it another way, each row and column has **an odd number of odd numbers** in it: either one or three odd numbers. There are five odd numbers in 1--9, so the only way they can distribute themselves, to get all row and column sums odd, is a row with three odd numbers and two rows with one odd number, and likewise for the columns. A row that has no odd numbers, or two of them, will add up to something even.

(paragraphs below edited 2017-06-09)

A second insight is that we can exchange any rows or any columns without affecting the row and column sums. So that means ("without loss of generality" as the mathematicians say) that we only need to consider arrays with a particular type of number, say an odd number, at the top left corner. Since there are only five odd numbers, if we have one of them top left, the other four odd numbers must also be in the first row and first column, with the four even numbers at the bottom right.

That means we have numbers of two classes, odd and even, with four of each. We also have two "boxes", box 1 being the first row and column, and box 2 being elsewhere. We need all four of the odd numbers to be in box 1 and all four of the even numbers to be in box 2. This is a hypergeometric probability:

$$
{ 4 \choose 4 } {4 \choose 0} \over {8 \choose 4} }
$$

which is

```{r}
1/choose(8,4)
```

or directly

```{r}
dhyper(4,4,4,4)
```

And then we have to recognize that the the top left number can be any one of the five odd numbers, so this answer needs to multiplied by 5:

```{r}
5*dhyper(4,4,4,4)
```

This is the same as my simulation gave, so I think it's correct.
