---
title: Summarizing columns in the tidyverse
author: Ken
date: '2017-07-14'
slug: summarizing-columns-in-the-tidyverse
categories: []
tags: []
---

## Introduction

I thought summarizing columns in the `tidyverse` was kind of clunky, at least until a couple of days ago. Let's read in some data to illustrate what I thought I had to do. This data comes from the Australian Institute for Sport; it's body measurements from male and female athletes who play ten different sports.

```{r}
library(tidyverse)
options(width=50)
my_url="http://www.utsc.utoronto.ca/~butler/c32/ais.txt"
athletes=read_tsv(my_url)
athletes
```

## What I thought I had to do

Let's suppose I want the mean and standard deviation of height and weight. It looked as if I had to do $2 \times 2=4$ different summarizations, like this:

```{r}
athletes %>% summarize(ht_mean=mean(Ht),
                       ht_sd=sd(Ht),
                       wt_mean=mean(Wt),
                       wt_sd=sd(Wt))
```

That seemed like a lot work, and rather repetitious. Couldn't I do better? Well, it turned out that I could, but I had some steps to go through first.

## Selecting columns

If you want to just display some columns of a data frame, `select` is very flexible. You can select columns by number:

```{r}
athletes %>% select(c(2,3))
```

or by names, *without* quotes:

```{r}
athletes %>% select(Sport:WCC)
```

But there is more. There's a whole family of [select helpers](http://dplyr.tidyverse.org/reference/select_helpers.html) that let you choose columns whose names start with something:

```{r}
athletes %>% select(starts_with("s"))
```

or end with something: `r tufte::margin_note("These are by default *not* case sensitive, but there is an option to make them so if you wish.")`

```{r}
athletes %>% select(ends_with("cc"))
```

or that contain something:

```{r}
athletes %>% select(contains("er"))
```

or that match a regular expression: `r tufte::margin_note("In this case, are exactly two characters long with the first one being H.")`

```{r}
athletes %>% select(matches("^H.$"))
```

You can also do things like selecting some or all of  variables with names like `x1`, `x2`, `x3`. There are many possibilities.

## Summarizing the columns you selected

So the selection is the first part of the story. The next question is, is there a nice way to summarize, in various different ways, those columns you have selected? There is, and it's called `summarize_at`. To go back to our original wish to get mean and SD of height and weight, note that `Ht` and `Wt` are the only two two-letter column names that end in T, so a regular expression will get them, and then this: `r tufte::margin_note("Inside summarize_at go two things, a vars inside which is something that will select the columns you want, and a funs inside which go the names of the summaries you want. These have to return single numbers.")`

```{r}
athletes %>% summarize_at(vars(matches("^.t$")),
                          funs(mean,sd))
```

And, as you see, we have the two means and the two SDs, without repeating ourselves.

The only restriction I see here is that if your summary returns more than one number, like a five-number summary:

```{r}
quantile(athletes$Ht)
```

then you have to resort to something like `map` and put the pieces together again:`r tufte::margin_note("Like Humpty Dumpty.")`

```{r}
athletes %>% select(matches("^.t$")) %>% map(quantile) %>% bind_rows()
```

This selects the height and weight columns, and for each of those, works out the five-number summary, and turns the results back from a list into a data frame. The downside to this is that the "names" attribute of the five-number-summary vectors gets lost, and you have to see that the first row is the minimum, the second row is the first quartile, and so on. 

## Summarizing columns that have a certain property

You might want to summarize all the columns that are numeric, or are dates, or something like that. For example, it only makes sense to calculate the mean of a column that is numeric. This kind of thing is done by `summarize_if`, which is like `summarize_at` except that the `vars` slot is replaced by something that will be true for all the columns you want, like this:

```{r}
athletes %>% summarize_if(is.numeric,funs(mean))
```

This finds the means of all the numeric columns. You'll need to pipe this into `print` to display the means that don't show here.

## By groups

The above summaries work for groups as well. To do that, insert a `group_by` before the `summarize_at`. For example, the mean and SD of height and weight for male and female athletes separately goes like this:

```{r}
athletes %>% group_by(Sex) %>% 
  summarize_at(vars(matches("^.t$")),
                    funs(mean,sd))
```


## Looking further

There are also `mutate_at` and `mutate_if`. These do calculations for columns whose names have something in common, or which have a certain property. This can be an efficient way to do "batch" transformations of variables, such as taking the logs of several variables, without the repetitiousness of transforming them one at a time.