---
title: How to change the significant digits in a tibble
author: Ken
date: '2018-07-13'
slug: how-to-change-the-significant-digits-in-a-tibble
categories:
  - R
tags:
  - rstats
  - tidyverse
---

## packages

```{r}
library(tidyverse)
library(lubridate)
```


## Introduction

When you display a tibble, or a data frame as a tibble, it is made to look nice and to show the most important digits. Let's illustrate with the inevitable `mtcars`:

```{r}
as_tibble(mtcars)
```

This shows, by default, three significant digits. Compare that with displaying it as a `data.frame` (which is what `mtcars`) actually is:

```{r}
mtcars
```

Usually, this is what you want. But not always. So how do you change it, if you want to? There's a simple answer, but I have a story to tell first.

## The Ice Classic

There is [a competition](http://www.nenanaakiceclassic.com/) each year in Nenana, Alaska, where people have to guess when the ice will melt, down to the nearest minute. Precisely, 
A wooden tripod is
placed on the frozen river, and people try to guess the exact minute
when the ice melts enough for the tripod to fall through the ice. The
contest started in 1917 as an amusement for railway workers, and has
taken place every year since. Now, hundreds of thousands of people
enter their guesses on the Internet and the prize for the winner can
be as much as $300,000. 

Because so much money is at stake, and because the exact same tripod
is placed at the exact same spot on the ice every year, the data are
consistent and accurate. `r tufte::margin_note("Does this sound like the preamble of an assignment question? It was.")`

Let's read in the data:

```{r}
myurl="http://www.utsc.utoronto.ca/~butler/c32/nenana.txt"
nenana=read_tsv(myurl)
nenana
```

The interesting thing here is `JulianDate`. This is the fractional number of days from January 1 when the ice melted, each year. Perhaps the most interesting thing to do with these data is to see whether there is a trend in this with `Year`:

```{r}
ggplot(nenana,aes(x=Year,y=JulianDate))+geom_point()+geom_smooth()
```

There is a lot of variability, but since about 1960 there appears to be a downward trend in `JulianDate`: that is, the ice is typically melting earlier each year.

I wanted to see whether the Julian date had been calculated correctly. In particular, there is a function `yday` in `lubridate` that calculates the number of the day within a year from a date, and since we have the actual dates we can check this. The first step is to turn the `Date&Time` into a proper date-time, and the first part of that step is to glue the `Year` onto the front of it. I use `str_c` from `stringr` for this:

```{r}
nenana %>% 
  mutate(long_date=str_c(Year," ",`Date&Time`)) %>% 
  mutate(dt=ymd_hm(long_date,tz="America/Anchorage")) %>% 
  select(-`Date&Time`)
```

That seems to have worked. Note that `lubridate` has handled the AM/PM properly.
Now I can calculate the Julian dates from `dt` and compare with `JulianDate` that came from the data file:

```{r}
nenana %>% 
  mutate(long_date=str_c(Year," ",`Date&Time`)) %>% 
  mutate(dt=ymd_hm(long_date,tz="America/Anchorage")) %>% 
  mutate(jd=yday(dt)) ->
nenana2 
nenana2 %>% 
  select(Year,JulianDate,jd)
```

Wait, some of these are not the same. Did the Ice Classic people make an error, or did I? (Place your bets now.)

There is a clue here, which is the `dbl` at the top of each column: these are decimal numbers that got rounded to 3 significant digits. What is it about the ones that got rounded wrong?

```{r}
nenana2 %>% 
  filter(round(JulianDate) != round(jd)) %>% 
  select(Year, JulianDate, jd, `Date&Time`)
```

These are all PMs, and the `JulianDate` in the data file is one more than the `jd` I calculated. You can certainly reason out what happened, but can we display a couple more decimal places in `JulianDate` and `jd`?

## Getting extra decimal places

When you display a `tibble`, the display is controlled by the package `pillar` (one of the many packages loaded by the `tidyverse`: the philosophy is that each aspect of the `tidyverse` is controlled by a separate relatively small package). So how do you control what `pillar` does? There is a variable `pillar.sigfigs` (that I had to do a fair amount of digging to find), which is set as an Option:

*** this doesn't actually display with more decimals

```{r}
options(pillar.sigfigs=100)
nenana2 %>% 
  filter(round(JulianDate) != round(jd)) %>% 
  select(JulianDate, jd
         ) %>% 
  print(width=80)
```

Now we have an answer to our question. `yday` calculates the *integer part* of the Julian date, but if the ice melted after noon, the fractional part of `JulianDate` is bigger than 0.5, and `JulianDate` will get rounded *up* to the next integer. That's why those `JulianDate` values appear to be 1 too big.

The solution involved thinking about what `yday` actually does, but we were deceived by the display of the tibble, and we had to realize that those apparent integers were actually rounded decimal numbers.


