---
title: Getting data from a frequency table, the tidyverse way
author: Ken
date: '2018-06-01'
slug: getting-data-from-a-frequency-table-the-tidyverse-way
categories:
  - statistics
tags:
  - rstats
  - tidyverse
---

## The usual package

```{r}
library(tidyverse)
```


## Introduction

`r tufte::margin_note("References at the end.")`

I was searching for some data for an assignment for my students, and I found an exercise in Watkins, Schaeffer and Cobb that seemed to fit my needs. However, the data values were in a frequency table, and I wanted an actual list of values, so that my students could make a histogram and calculate some summary statistics using R. Also, for reasons that we will come to later, I wanted to randomly shuffle the values, which turns out to have a ludicrously easy solution in the Tidyverse.

## The data set, as it reached me

These are speeds (in miles per hour) of 1000 cars measured on a particular stretch of road (in free-flowing traffic).

Here's the frequency table:

```{r}
counts=tribble(
  ~speed_int, ~count,
  25, 2, 
  26, 31,
  27, 92,
  28, 149,
  29, 178,
  30, 156,
  31, 157,
  32, 99,
  33, 74,
  34, 31,
  35, 16,
  36, 13,
  37, 1, 
  38, 1
)
counts
```

The speeds are in `speed_int` and the `count` column contains frequencies.

What I want is a list of speeds, each speed appearing in the data set the corresponding `count` number of times. 
That is, the frequencies add up to 1000, so I want 1000 values. 


## Repeating the values the right number of times

Base R's `rep` function provides the simplest way to get what I want:

```{r}
with(counts,rep(speed_int,count))
```

How do I do the same thing the tidyverse way? The obvious idea doesn't work:

```{r, eval=F}
counts %>% mutate(speed=rep(speed_int,count))
```

This is because the new `speed` column has to have the same length as the old `speed_int` column, and it's now longer. 
So I thought a list-column might do the job: replicate each `speed_int` by the count in the corresponding row, 
and save each vector answer in a list-column, thus:

```{r}
counts %>% mutate(speed=map2(speed_int,count,~rep(.x,.y)))
```

I don't really need the `~` form here, but I think it reads nicely: "for each `speed_int` and `count` (in parallel), 
repeat the first thing as many times as the second thing". Each entry in `speed` is a vector as long as the entry in `count`. 

## Making a data frame of speeds

Now, I need to get the actual speeds out of there. This is a job for `unnest`:

```{r}
counts %>% mutate(speed=map2(speed_int,count,~rep(.x,.y))) %>% 
  unnest()
```

25 appears twice, then 26 31 times, and so on.

Now it occurs to me that the values in `speed_int` have been repeated the right number of times, 
so I didn't need to work so hard: *anything* that produced a vector of the right length would have done. 
But anyway.

## Shuffling the rows

I wanted to present my students with a file containing a bunch of speeds in apparently random order 
(as if they were the speeds of 1000 cars in the order they passed the speed gun). I remembered `sample_frac`, 
which randomly selects a fraction of the rows of a data frame, but didn't realize (until I looked at the Stack Overflow post
referenced below) that sampling a fraction 1 of the rows will sample them all, but in a shuffled order, in the same
way that `sample` when not supplied a sample size will randomly sample all of its input (that is, shuffle it). 
Thus I can do this:

```{r}
counts %>% mutate(speed=map2(speed_int,count,~rep(.x,.y))) %>% 
  unnest() %>% 
  sample_frac(1)
```

and do a bit of tidying before I save the results:

```{r}
counts %>% mutate(speed=map2(speed_int,count,~rep(.x,.y))) %>% 
  unnest() %>% 
  sample_frac(1) %>% 
  mutate(car_no=row_number()) %>% 
  select(car_no,speed)
```

The last two lines create a column of "car numbers", and then put the two columns I want to keep in a sensible order.


## References

[Using `rep`](https://stats.stackexchange.com/questions/15574/how-to-convert-a-frequency-table-into-a-vector-of-values?utm_medium=organic&utm_source=google_rich_qa&utm_campaign=google_rich_qa)

[Randomly shuffling rows of a data frame](https://stackoverflow.com/questions/6422273/how-to-randomize-or-permute-a-dataframe-rowwise-and-columnwise?utm_medium=organic&utm_source=google_rich_qa&utm_campaign=google_rich_qa)

[Watkins, Schaeffer and Cobb](https://www.amazon.ca/Statistics-Binder-Ready-Version-Decision/dp/0470559942/ref=sr_1_1?ie=UTF8&qid=1527868509&sr=8-1&keywords=watkins+scheaffer+cobb+statistics&dpID=51kUOPO2Y0L&preST=_SX198_BO1,204,203,200_QL40_&dpSrc=srch)