---
title: Un-counting
author: Ken
date: '2019-07-13'
slug: un-counting
categories:
  - stats
tags:
  - '#rstats'
  - '#tidyverse'
---

## Packages

```{r}
library(tidyverse)
```

## Introduction

You probably know about `count`, which tells you how many observations you have in each group:

```{r}
d <- tribble(
  ~g, ~y,
  "a", 10,
  "a", 13,
  "a", 14, 
  "a", 14,
  "b", 6,
  "b", 7,
  "b", 9
)
```

There are four observations in group `a` and three in group `b`:

```{r}
d %>% count(g) -> counts
counts
```

I didn't know about this until fairly recently. Until then, I thought you had to do this:

```{r}
d %>% group_by(g) %>% 
  summarize(count=n()) 
```

which works, but is a lot more typing.

## Going the other way

The other day, I had the opposite problem. I had a table of frequencies, and I wanted to get it back to one row per observation (I was fitting a model in Stan, and I didn't know how to deal with frequencies). I had no idea how you might do that (without something ugly like loops), and I was almost embarrassed to stumble upon this:

```{r}
counts %>% uncount(n)
```

My situation was a bit less trivial than that. I had disease category counts of coal miners with different exposures to coal dust:

```{r}
my_url="https://www.utsc.utoronto.ca/~butler/d29/miners-tab.txt"
miners0 <- read_table(my_url)
miners0
```

This needs tidying to get the frequencies all into one column:

```{r}
miners0 %>% 
  gather(disease, freq, -Exposure) -> miners
miners
```

So I wanted to fit an ordered logistic regression in Stan, predicting disease category from exposure, but I didn't know how to handle the frequencies. If I had one row per miner, I thought...

```{r}
miners %>% uncount(freq) 
```

... and so I do. (I scrolled down to check, and *eventually* got past the 98 miners with 5.8 years of exposure and no disease).

From there, you can use [this](https://mc-stan.org/docs/2_19/stan-users-guide/ordered-logistic-section.html) to fit the model, though I would rather have weakly informative priors for their `beta` and `c`. `c` is tricky, since it is ordered, but I used the idea [here](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations) (near the bottom) and it worked smoothly.

## An apology

I know `utteranc.es` appears down below, but I haven't been able to make it work yet. If you want to comment, [find my announcement](https://twitter.com/KenButler12/status/1150228575653695488) of this post on Twitter and comment there, if you would be so kind.